The \tsc{r}-pronoun and postposition \tit{waar-mee} in Dutch

### 2020-01-16 paper.tex
leaving jan's place
refereces in place

\title{The \tsc{r}-pronoun and postposition \tit{waarmee} in Dutch}
\author{Fenna Bergsma}
\date{\today}

@@

This chapter is an in-depth study of the instrumental \tsc{r}-pronoun and postposition \tit{'r -mee} `with it' and \tit{waarmee} `with what' in Dutch. This instance is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the \tsc{r}-pronoun is syncretic with the locative. In many of the \tsc{r}-pronouns and postpositions, the meaning component of the locative is intuitive, as many prepositions express locations, directions etc. However, an instrumental expresses an instrument, which does not have a meaning component assosicated with location. Ideally, an analysis treats the syncretism not as accidental but allows for the locative meaning to be absent.
The second reason why I focus on this particular \tsc{r}-pronoun and postposition has to do with the form of the adposition. The preposition \tit{met} `with' does not only turn into a postposition, but it also changes into \tit{-mee} `with' when it is combined with an \tsc{r}-pronoun. This last observervation has so far remained unexplained.

The main generalization is that the instrumental \tsc{r}-pronoun and postposition \tit{waarmee} `with what' takes precedence over the instrumental preposition and inanimate pronoun \tit{met wat} `with what'. This generalization is subject to an important condition: the instrumental object needs to form a proper constituent i.e. a constituent to the exclusion of other features. When this condition is not met, the preposition and pronoun appear. This can straightforwardly follow in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. I work this idea out capturing the following observations. First, \tsc{r}-pronouns are syncretic with locatives. Second, regular pronouns appear with prepoositions, \tsc{r}-pronouns with postpositions. Third, the instrumental preposition and postposition differ in form (\tit{met} vs. \tit{mee}).

This paper is structured as follows. In Section \ref{sec:distribution} I discuss the condition that \tsc{r}-pronouns and postpositions are subject too: they need to form a proper constituent. I show that \tsc{r}-pronouns and postpositions and prepositions and pronouns are in complementary distribution, depending on whether all features form a proper constituent or not.
In Section \ref{sec:takingapart} I decompose \tit{waarmee} `with what' and \tit{met wat} `with what', and I connect them to parts of syntactic structure. Section \ref{sec:derivation} shows in a derivation how \tit{waarmee} `with what' is derived and the role that constituency plays. Section \ref{sec:conclusion} concludes and discusses possible extensions of the analysis. All examples in this paper are from Dutch, unless indicated otherwise. Unmarked examples are constructed and have been verified by native speakers.

@@

The goal of this section is to show that \tsc{r}-pronouns and postpositions appear when all features that are to be spelled out form a proper constituent, i.e. that all features form a constituent to the exclusion of any other features. I start the section by showing that \tit{'rmee} `with it' and \tit{waarmee} `with what` are the default, that normally they appear instead of \tit{met 't} `with it' and \tit{met wat} `with what`. Then I show an exception, in which \tit{met wat} `with what` has to be used and \tit{waarmee} `with what` cannot. I argue that the crucial difference between the exception and other cases is that in the exception the features do not form a proper constituent.

@@

\ex. \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder 'r mee.\\
 I paint there -with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee 'r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:erprep}

@@

The situation of the inanimate wh-pronouns resembles the one of the inanimate personal pronouns. \tit{Wat} `what' can function as an object of a verb (see \ref{ex:wat}), but not as an object of a preposition \ref{ex:metwat}. In that case, the \tsc{r}-pronoun and postposition \tit{waarmee} `with what' appears, as shown in \ref{ex:waarmee}.\footnote{The sentence in \ref{ex:metwat} is unacceptable with neutral intonation. It becomes acceptable if \tit{wat} `what' is stressed, for example in a context in which the speaker is highly surprised about the choice of the object the hearer is painting with.}

\ex.
\ag. Wat zie jij?\\
 what see you\\
 `What do you see?'\label{ex:wat}
\bg. *Met wat schilder jij?\\
 with what paint you\\
 `What are you painting with?'\label{ex:metwat}
\bg. Waarmee schilder jij?\\
 {where with} paint you with\\
 `What are you painting with?'\label{ex:waarmee}

\tit{Waarmee} `with what' and not \tit{met wat} `with what' does not only appear in wh-questions, but also in other contexts. \ref{ex:headed} gives an example of a headed relative, and \ref{ex:headless} shows a free relative in which both predicates combine with an instrumental object. The use of \tit{met wat} `with what' is ungrammatical in both contexts, and \tit{waarmee} `with what' is used.

\ex.\label{ex:headed}
\ag. Ik schilder met de kwast waarmee jij ook schildert.\\
 I paint with the brush {where with} you also paint\\
 `I am painting with the brush that you are painting with too.'
\bg. *Ik schilder met de kwast met wat jij ook schildert.\\
 I paint with the brush with what you also paint\\
 `I am painting with the brush that you are painting with too.'

 \ex.\label{ex:headless}
 \ag. Ik schilder waarmee jij ook schildert.\\
  I paint {where with} you also paint\\
  `I am painting with what you are painting with too.'
 \bg. *Ik schilder met wat jij ook schildert.\\
  I paint with what you also paint\\
  `I am painting with what you are painting with too.'

@@

This section discusses an exception to what is shown in the previous section. I show an instance in which \tit{met wat} `with what' has to be used instead of \tit{waarmee} `with what'. This instance comes from a mismatching free relative construction.

A mismatching free relative is a free relative construction in which the two predicates (the one in the main clause and the one in the embedded clause) combine with two different cases (i.e. the case requirements do not match). I give an example in which the \tsc{r}-pronoun and postposition appears before I get to the exception. Consider \ref{ex:gekochtwaarmee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental object. The predicate in the main clause clause, \tit{gekocht} `bought' combines with an accusative object. The \tsc{r}-pronoun and postposition \tit{waarmee} `with what' is used here. The use of \tit{met wat} `with what' is ungrammatical in this context, illustrated in \ref{ex:gekochtmetwat}.\footnote{In this example, \tit{waar} `where' takes \tit{-mee} `with' to the left edge of the embedded clause. It is also possible for \tit{-mee} `with' to be stranded, and \tit{waar} `where' to be moved to the left edge of the embedded clause on its own.

@@

\ex.
\ag. Ik heb gekocht waarmee jij schildert.\\
 I have bought {where with} you paint\\
 `I bought what you are painting with.'\label{ex:gekochtwaarmee}
\bg. *Ik heb gekocht met wat jij schildert.\\
 I have bought with what you paint\\
 `I bought what you are painting with.'\label{ex:gekochtmetwat}

If the predicates are switched around between the clauses, the \tsc{r}-pronoun and postposition do not appear anymore. In \ref{ex:schildermet}, \tit{schilder} `paint' combines with an instrumental object in the main clause and \tit{gekocht} `bought' combines with an accusative object in the embedded clause. The use of an \tsc{r}-pronoun and postposition is ungrammatical, as indicated by \ref{ex:schilderwaarmee}. Instead, a combination of the regular instrumental preposition \tit{met} `with' and the regular wh-pronoun \tit{wat} `what' in used.

\ex.\label{ex:schildermet}
\ag. *Ik schilder waarmee jij hebt gekocht.\\
 I paint {where with} you have bought\\
 `I paint with what you bought.'\label{ex:schilderwaarmee}
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:schildermetwat}

Table \ref{tbl:distribution} summarizes the pattern. When the main clause predicate combines with an accusative and the embedded clause predicate with an instrumental, \tit{waarmee} `with what' is grammatical and \tit{met wat} `with what' is ungrammatical. When the main clause predicate combines with an instrumental and the embedded clause predicate with an accusative, \tit{waarmee} `with what' is ungrammatical and \tit{met wat} `with what' is used.

\begin{table}[ht]
	\center
	\caption {Distribution between \tit{waarmee} and \tit{met wat}}
	\begin{minipage}{0.45\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                              & \tit{waarmee} & \tit{met wat} \\
		\midrule
    m:\tsc{acc}, e:\tsc{ins}  & ✔             & *             \\
    m:\tsc{ins}, e:\tsc{acc}  & *             & ✔             \\
    \bottomrule
\end{tabularx}
\label{tbl:distribution}
\end{minipage}
\end{table}

@@

\ex.
\ag. Ik schilder 'r mee.\\
 I paint there with\\
 `I am painting with it.'\label{ex:const1}
\bg. Waarmee schilder jij?\\
{where with} paint you with\\
 `What are you painting with?'\label{ex:const2}
\bg. Ik schilder met de kwast [waarmee jij ook schildert].\\
 I paint with the brush {where with} you also paint\\
 `I am painting with the brush that you are painting with too.'\label{ex:const3}
\bg. Ik schilder [waarmee jij ook schildert].\\
 I paint {where with} you also paint\\
 `I am painting with what you are painting with too.'\label{ex:const4}

@@

\ex.
\a. [[ik] [[schilder] ['r mee]]]\label{ex:const1stage}
\b. [[jij] [[schilder] [waarmee]]]\label{ex:const2stage}
\b. [[jij] [[ook] [[schilder] [waarmee]]]]\label{ex:const3stage}

@@

I repeat the mismatching free relative in which \tit{waarmee} `with what' appears in \ref{ex:mismatchwaarmee}. The predicate \tit{schildert} `paints' combines in the embedded clause with the instrumental object. The instrumental object forms a proper constituent within the embedded clause, and the it can be realized as the \tsc{r}-pronoun and postposition \tit{waarmee} `with what'.\footnote{I assume that the accusative case requirement of \tit{gekocht} `bought' is satisfying by grafting a subconstituent of \tit{waarmee} `with what' \citep{bergsma2019}.}

\exg. Ik heb gekocht [waarmee jij schildert].\\
 I have bought {where with} you paint\\
 `I bought what you are painting with.'\label{ex:mismatchwaarmee}

Next, we arrive at the mismatching free relative in which \tit{waarmee} `with what' cannot be used, but \tit{met wat} `with what' appears. The embedded clause predicate \tit{gekocht} `bought' combines with an accusative DP. The accusative object of a verb is always \tit{wat} `what', as I showed in \ref{ex:wat}. The instrumental only comes into the picture in the main clause, when \tit{schilder} `paint' combines with an instrumental object. At no point in the derivation does the instrumental object form a proper constituent, and \tit{waarmee} `with what' does not surface.

@@

 The mismatching free relative in \ref{ex:grammaticalwat} is not the only construction in which the string \tit{met wat} `with what' appears. I give examples of two more occurrences in \ref{ex:moremetwat}. In \ref{ex:watindef}, \tit{wat} `what' is the \tit{wat} `what' in the so-called \tit{wat voor} `what for'-construction \citep[cf.][]{corver1991}.
 In \ref{ex:watindef}, \tit{wat} appears as a quantifier, and it means `some'. In both constructions \tit{wat} `what' takes a complement and \tit{met wat} `with what' do not form a proper constituent. The brackets within the examples indicate the constituency.

@@

\ref{ex:summaryconst} summarizes what I showed in this section. \tit{Met wat} `with what' can never surface when \tit{met} `with' and \tit{wat} `what' form a proper constituent. It always becomes \tit{waarmee} `with what', as shown in \ref{ex:waarmeefr}.

\ex.\label{ex:summaryconst}
\a. [[met] [wat]] → [waarmee]\label{ex:waarmeefr}
\b. [met [wat [X]]]\label{ex:metwatx}
\b. [met [[wat] [X]]]\label{ex:metwatfr}

@@

\section{Decomposing \tit{waarmee} and \tit{met wat}}\label{sec:takingapart}

In this section I decompose \tit{waarmee} `with what' and \tit{met wat} `met wat' into smaller units, and I connect them to parts of syntactic structure. Both spell out the same set of features but the distribution is different. I decompose \tit{waarmee} and \tit{met wat} as I show in \ref{ex:decompose}. Earlier works have decomposed these pronouns in similar ways (cf. \citet{hachem2015,noonan2017dutch,wesseling2018}).

@@

Let me start with the morphemes \tit{w} and \tit{a} that appear in both expressions. I assume that they correspond to the same syntactic structure in both \tit{waarmee} `with what' and \tit{met wat} `with what'. As I am interested in the differencs between the two expression, I do not discuss the featural content of \tit{w} and \tit{a} into depth.

@@

I put \tit{w} and \tit{a} aside for now, assuming they spell out the same syntactic structure in \tit{waarmee} `with what' and \tit{met wat} `with what'. This leaves \tit{'r -mee} `with it' and \tit{met 't} `with it'.

@@

In the next section I put all features back together in a derivation and I show how \tit{waarmee} `with what' surfaces when all features form a constituent. \tit{Met wat} `with what' appears when the functional sequence is disrupted.

@@

Before I show that \tit{waarmee} `with what' is used when all features form a proper constituent, I need to make some assumptions about the spellout process in Nanosyntax explicit. Spellout happens in a cyclic derivation, following a spellout algorithm \citep{starke2018}. After each instance of merge, spellout takes place. If no spellout exists for the phrase created by the newly added feature, evacuation movements specified in the spellout algorithm take place. The algorithm is given in \ref{ex:spellout}.

@@

I skip over the details of how \tit{w} and \tit{a} end up in their positions.\footnote{I assume that \tsc{wP} and \tsc{deixP} are both complex specifiers that are created after \tsc{thing} is spelled out in \ref{ex:thingspellout}. After each instance of merge after that, backtracking takes place, the complex specifier is detached from the structure and the case features are spelled out together with or as a postposition on \tsc{thing}.} The final result of the structure for \tit{waarmee} `with what' look as in \ref{ex:spelloutwaarmee}.

\ex. \begin{forest} boom
[\tsc{\tsc{wP}}
    [\tsc{\tsc{wP}}
        [\tsc{w}, roof]
    ]
    {\draw (.east) node[right]{⇒ \tit{w}}; }
    [\tsc{deixP}
        [\tsc{deixP}
            [\tsc{deix}, roof]
        ]
        {\draw (.east) node[right]{⇒ \tit{a}}; }
        [\tsc{insP}
            [\tsc{thingP}
               [\tsc{thing}, roof]
            ]
            {\draw (.east) node[right]{⇒ \tit{'r}}; }
            [\tsc{insP}
               [\tsc{f4}]
               [\tsc{datP}
                   [\tsc{f3}]
                   [\tsc{accP}
                       [\tsc{f2}]
                       [\tsc{nomP}
                           [\tsc{f1}]
                           [ΣP
                               [Σ]
                           ]
                       ]
                   ]
               ]
            ]
            {\draw (.east) node[right]{⇒ \tit{-mee}}; }
        ]
    ]
]
\end{forest}\label{ex:spelloutwaarmee}

@@

There is no constituent in \ref{ex:spelloutwaarmee} that contains \tit{waar} but not \tit{-mee}. To resolve this situation I follow \citet{noonan2017dutch} in assuming that the phrase containing the adposition (here \tit{-mee}) can syntactically move to a position higher in the structure. Evidence that \tit{mee} `with' has moved comes from the ungrammaticality of \ref{ex:meeend}. This shows that \tit{mee} `with' cannot be stranded in its base position (see \ref{ex:const3stage}).

@@

So far I showed how \tit{waarmee} `with what' is derived if all syntactic features form a constituent. Next I address how \tit{waarmee} `with what' is blocked and \tit{met wat} `with what' appears when the features do not form a proper constituent is derived. An example of a situation in which all features do not form a constituent is given in \ref{ex:noconst}.

@@

This section showed how the instrumental inanimate relative pronoun is realized as \tit{waarmee} `with what' when all syntactic features form a constituent. It also showed how \tit{met wat} `with what' appears when all features do not form a proper constituent, and other features intervene.

@@

In this paper, I discussed the instrumental \tsc{r}-pronoun and postposition \tit{waarmee} `with what' in Dutch. The main intuition is that this form that appears when features of an instrumental inanimate object form a proper constituent. The form cannot appear, and \tit{met wat} `with what` surfaces instead, when all features do not form a proper constituent. The described pattern follows from a core assumption in Nanosyntax: phrasal spellout spells out constituents \citep{starke2009}. The distribution of \tit{r}-pronouns and postpositions and prepositions and pronouns follows from regular spellout mechanisms.

>>>>

\title{The \tsc{r}-pronoun and postposition \tit{waar-mee} in Dutch}
\author{Fenna Bergsma}
\date{\today}

@@

This chapter is an in-depth study of the instrumental \tsc{r}-pronoun and postposition \tit{'r-mee} `with it' and \tit{waar-mee} `with what' in Dutch. This instance is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the \tsc{r}-pronoun is syncretic with the locative. In many of the \tsc{r}-pronouns and postpositions, the meaning component of the locative is intuitive, as many prepositions express locations, directions etc. However, an instrumental expresses an instrument, which does not have a meaning component assosicated with location. Ideally, an analysis treats the syncretism not as accidental but allows for the locative meaning to be absent.
The second reason why I focus on this particular \tsc{r}-pronoun and postposition has to do with the form of the adposition. The preposition \tit{met} `with' does not only turn into a postposition, but it also changes into \tit{-mee} `with' when it is combined with an \tsc{r}-pronoun. This last observervation has so far remained unexplained.

The main generalization is that the instrumental \tsc{r}-pronoun and postposition \tit{waar-mee} `with what' takes precedence over the instrumental preposition and inanimate pronoun \tit{met wat} `with what'. This generalization is subject to an important condition: the instrumental object needs to form a proper constituent i.e. a constituent to the exclusion of other features. When this condition is not met, the preposition and pronoun appear. This can straightforwardly follow in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. I work this idea out capturing the following observations. First, \tsc{r}-pronouns are syncretic with locatives. Second, regular pronouns appear with prepoositions, \tsc{r}-pronouns with postpositions. Third, the instrumental preposition and postposition differ in form (\tit{met} vs. \tit{mee}).

This paper is structured as follows. In Section \ref{sec:distribution} I discuss the condition that \tsc{r}-pronouns and postpositions are subject too: they need to form a proper constituent. I show that \tsc{r}-pronouns and postpositions and prepositions and pronouns are in complementary distribution, depending on whether all features form a proper constituent or not.
In Section \ref{sec:takingapart} I decompose \tit{waar-mee} `with what' and \tit{met wat} `with what', and I connect them to parts of syntactic structure. Section \ref{sec:derivation} shows in a derivation how \tit{waar-mee} `with what' is derived and the role that constituency plays. Section \ref{sec:conclusion} concludes and discusses possible extensions of the analysis. All examples in this paper are from Dutch, unless indicated otherwise. Unmarked examples are constructed and have been verified by native speakers.

@@

The goal of this section is to show that \tsc{r}-pronouns and postpositions appear when all features that are to be spelled out form a proper constituent, i.e. that all features form a constituent to the exclusion of any other features. I start the section by showing that \tit{'rmee} `with it' and \tit{waar-mee} `with what` are the default, that normally they appear instead of \tit{met 't} `with it' and \tit{met wat} `with what`. Then I show an exception, in which \tit{met wat} `with what` has to be used and \tit{waar-mee} `with what` cannot. I argue that the crucial difference between the exception and other cases is that in the exception the features do not form a proper constituent.

@@

\ex. \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder 'r -mee.\\
 I paint there -with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee 'r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:erprep}

@@

The situation of the inanimate wh-pronouns resembles the one of the inanimate personal pronouns. \tit{Wat} `what' can function as an object of a verb (see \ref{ex:wat}), but not as an object of a preposition \ref{ex:metwat}. In that case, the \tsc{r}-pronoun and postposition \tit{waar-mee} `with what' appears, as shown in \ref{ex:waar-mee}.\footnote{The sentence in \ref{ex:metwat} is unacceptable with neutral intonation. It becomes acceptable if \tit{wat} `what' is stressed, for example in a context in which the speaker is highly surprised about the choice of the object the hearer is painting with.}

\ex.
\ag. Wat zie jij?\\
 what see you\\
 `What do you see?'\label{ex:wat}
\bg. *Met wat schilder jij?\\
 with what paint you\\
 `What are you painting with?'\label{ex:metwat}
\bg. Waar -mee schilder jij?\\
 where with paint you with\\
 `What are you painting with?'\label{ex:waar-mee}

\tit{Waar-mee} `with what' and not \tit{met wat} `with what' does not only appear in wh-questions, but also in other contexts. \ref{ex:headed} gives an example of a headed relative, and \ref{ex:headless} shows a free relative in which both predicates combine with an instrumental object. The use of \tit{met wat} `with what' is ungrammatical in both contexts, and \tit{waar-mee} `with what' is used.

\ex.\label{ex:headed}
\ag. Ik schilder met de kwast waar -mee jij ook schildert.\\
 I paint with the brush where with you also paint\\
 `I am painting with the brush that you are painting with too.'
\bg. *Ik schilder met de kwast met wat jij ook schildert.\\
 I paint with the brush with what you also paint\\
 `I am painting with the brush that you are painting with too.'

 \ex.\label{ex:headless}
 \ag. Ik schilder waar -mee jij ook schildert.\\
  I paint where with you also paint\\
  `I am painting with what you are painting with too.'
 \bg. *Ik schilder met wat jij ook schildert.\\
  I paint with what you also paint\\
  `I am painting with what you are painting with too.'

@@

This section discusses an exception to what is shown in the previous section. I show an instance in which \tit{met wat} `with what' has to be used instead of \tit{waar-mee} `with what'. This instance comes from a mismatching free relative construction.

A mismatching free relative is a free relative construction in which the two predicates (the one in the main clause and the one in the embedded clause) combine with two different cases (i.e. the case requirements do not match). I give an example in which the \tsc{r}-pronoun and postposition appears before I get to the exception. Consider \ref{ex:gekochtwaar-mee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental object. The predicate in the main clause clause, \tit{gekocht} `bought' combines with an accusative object. The \tsc{r}-pronoun and postposition \tit{waar-mee} `with what' is used here. The use of \tit{met wat} `with what' is ungrammatical in this context, illustrated in \ref{ex:gekochtmetwat}.\footnote{In this example, \tit{waar} `where' takes \tit{-mee} `with' to the left edge of the embedded clause. It is also possible for \tit{-mee} `with' to be stranded, and \tit{waar} `where' to be moved to the left edge of the embedded clause on its own.

@@

\ex.
\ag. Ik heb gekocht waar -mee jij schildert.\\
 I have bought where with you paint\\
 `I bought what you are painting with.'\label{ex:gekochtwaar-mee}
\bg. *Ik heb gekocht met wat jij schildert.\\
 I have bought with what you paint\\
 `I bought what you are painting with.'\label{ex:gekochtmetwat}

If the predicates are switched around between the clauses, the \tsc{r}-pronoun and postposition do not appear anymore. In \ref{ex:schildermet}, \tit{schilder} `paint' combines with an instrumental object in the main clause and \tit{gekocht} `bought' combines with an accusative object in the embedded clause. The use of an \tsc{r}-pronoun and postposition is ungrammatical, as indicated by \ref{ex:schilderwaar-mee}. Instead, a combination of the regular instrumental preposition \tit{met} `with' and the regular wh-pronoun \tit{wat} `what' in used.

\ex.\label{ex:schildermet}
\ag. *Ik schilder waar -mee jij hebt gekocht.\\
 I paint where with you have bought\\
 `I paint with what you bought.'\label{ex:schilderwaar-mee}
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:schildermetwat}

Table \ref{tbl:distribution} summarizes the pattern. When the main clause predicate combines with an accusative and the embedded clause predicate with an instrumental, \tit{waar-mee} `with what' is grammatical and \tit{met wat} `with what' is ungrammatical. When the main clause predicate combines with an instrumental and the embedded clause predicate with an accusative, \tit{waar-mee} `with what' is ungrammatical and \tit{met wat} `with what' is used.

\begin{table}[ht]
	\center
	\caption {Distribution between \tit{waar-mee} and \tit{met wat}}
	\begin{minipage}{0.45\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                              & \tit{waar-mee} & \tit{met wat} \\
		\midrule
    m:\tsc{acc}, e:\tsc{ins}  & ✔             & *             \\
    m:\tsc{ins}, e:\tsc{acc}  & *             & ✔             \\
    \bottomrule
\end{tabularx}
\label{tbl:distribution}
\end{minipage}
\end{table}

@@

\ex.
\ag. Ik schilder 'r -mee.\\
 I paint there with\\
 `I am painting with it.'\label{ex:const1}
\bg. Waar -mee schilder jij?\\
where with paint you with\\
 `What are you painting with?'\label{ex:const2}
\bg. Ik schilder met de kwast [waar -mee jij ook schildert].\\
 I paint with the brush where with you also paint\\
 `I am painting with the brush that you are painting with too.'\label{ex:const3}
\bg. Ik schilder [waar -mee jij ook schildert].\\
 I paint where with you also paint\\
 `I am painting with what you are painting with too.'\label{ex:const4}

@@

\ex.
\a. [[ik] [[schilder] ['r -mee]]]\label{ex:const1stage}
\b. [[jij] [[schilder] [waar-mee]]]\label{ex:const2stage}
\b. [[jij] [[ook] [[schilder] [waar-mee]]]]\label{ex:const3stage}

@@

I repeat the mismatching free relative in which \tit{waar-mee} `with what' appears in \ref{ex:mismatchwaar-mee}. The predicate \tit{schildert} `paints' combines in the embedded clause with the instrumental object. The instrumental object forms a proper constituent within the embedded clause, and the it can be realized as the \tsc{r}-pronoun and postposition \tit{waar-mee} `with what'.\footnote{I assume that the accusative case requirement of \tit{gekocht} `bought' is satisfying by grafting a subconstituent of \tit{waar-mee} `with what' \citep{bergsma2019}.}

\exg. Ik heb gekocht [waar -mee jij schildert].\\
 I have bought where with you paint\\
 `I bought what you are painting with.'\label{ex:mismatchwaar-mee}

Next, we arrive at the mismatching free relative in which \tit{waar-mee} `with what' cannot be used, but \tit{met wat} `with what' appears. The embedded clause predicate \tit{gekocht} `bought' combines with an accusative DP. The accusative object of a verb is always \tit{wat} `what', as I showed in \ref{ex:wat}. The instrumental only comes into the picture in the main clause, when \tit{schilder} `paint' combines with an instrumental object. At no point in the derivation does the instrumental object form a proper constituent, and \tit{waar-mee} `with what' does not surface.

@@

 The mismatching free relative in \ref{ex:mismatchmetwat} is not the only construction in which the string \tit{met wat} `with what' appears. I give examples of two more occurrences in \ref{ex:moremetwat}. In \ref{ex:watindef}, \tit{wat} `what' is the \tit{wat} `what' in the so-called \tit{wat voor} `what for'-construction \citep[cf.][]{corver1991}.
 In \ref{ex:watindef}, \tit{wat} appears as a quantifier, and it means `some'. In both constructions \tit{wat} `what' takes a complement and \tit{met wat} `with what' do not form a proper constituent. The brackets within the examples indicate the constituency.

@@

\ref{ex:summaryconst} summarizes what I showed in this section. \tit{Met wat} `with what' can never surface when \tit{met} `with' and \tit{wat} `what' form a proper constituent. It always becomes \tit{waar-mee} `with what', as shown in \ref{ex:waar-meefr}.

\ex.\label{ex:summaryconst}
\a. [[met] [wat]] → [waar-mee]\label{ex:waar-meefr}
\b. [met [wat [X]]]\label{ex:metwatx}
\b. [met [[wat] [X]]]\label{ex:metwatfr}

@@

\section{Decomposing \tit{waar-mee} and \tit{met wat}}\label{sec:takingapart}

In this section I decompose \tit{waar-mee} `with what' and \tit{met wat} `met wat' into smaller units, and I connect them to parts of syntactic structure. Both spell out the same set of features but the distribution is different. I decompose \tit{waar-mee} and \tit{met wat} as I show in \ref{ex:decompose}. Earlier works have decomposed these pronouns in similar ways (cf. \citet{hachem2015,noonan2017dutch,wesseling2018}).

@@

Let me start with the morphemes \tit{w} and \tit{a} that appear in both expressions. I assume that they correspond to the same syntactic structure in both \tit{waar-mee} `with what' and \tit{met wat} `with what'. As I am interested in the differencs between the two expression, I do not discuss the featural content of \tit{w} and \tit{a} into depth.

@@

I put \tit{w} and \tit{a} aside for now, assuming they spell out the same syntactic structure in \tit{waar-mee} `with what' and \tit{met wat} `with what'. This leaves \tit{'r-mee} `with it' and \tit{met 't} `with it'.

@@

In the next section I put all features back together in a derivation and I show how \tit{waar-mee} `with what' surfaces when all features form a constituent. \tit{Met wat} `with what' appears when the functional sequence is disrupted.

@@

Before I show that \tit{waar-mee} `with what' is used when all features form a proper constituent, I need to make some assumptions about the spellout process in Nanosyntax explicit. Spellout happens in a cyclic derivation, following a spellout algorithm \citep{starke2018}. After each instance of merge, spellout takes place. If no spellout exists for the phrase created by the newly added feature, evacuation movements specified in the spellout algorithm take place. The algorithm is given in \ref{ex:spellout}.

@@

I skip over the details of how \tit{w} and \tit{a} end up in their positions.\footnote{I assume that \tsc{wP} and \tsc{deixP} are both complex specifiers that are created after \tsc{thing} is spelled out in \ref{ex:thingspellout}. After each instance of merge after that, backtracking takes place, the complex specifier is detached from the structure and the case features are spelled out together with or as a postposition on \tsc{thing}.} The final result of the structure for \tit{waar-mee} `with what' look as in \ref{ex:spelloutwaar-mee}.

\ex. \begin{forest} boom
[\tsc{\tsc{wP}}
    [\tsc{\tsc{wP}}
        [\tsc{w}, roof]
    ]
    {\draw (.east) node[right]{⇒ \tit{w}}; }
    [\tsc{deixP}
        [\tsc{deixP}
            [\tsc{deix}, roof]
        ]
        {\draw (.east) node[right]{⇒ \tit{a}}; }
        [\tsc{insP}
            [\tsc{thingP}
               [\tsc{thing}, roof]
            ]
            {\draw (.east) node[right]{⇒ \tit{'r}}; }
            [\tsc{insP}
               [\tsc{f4}]
               [\tsc{datP}
                   [\tsc{f3}]
                   [\tsc{accP}
                       [\tsc{f2}]
                       [\tsc{nomP}
                           [\tsc{f1}]
                           [ΣP
                               [Σ]
                           ]
                       ]
                   ]
               ]
            ]
            {\draw (.east) node[right]{⇒ \tit{-mee}}; }
        ]
    ]
]
\end{forest}\label{ex:spelloutwaar-mee}

@@

There is no constituent in \ref{ex:spelloutwaar-mee} that contains \tit{waar} but not \tit{-mee}. To resolve this situation I follow \citet{noonan2017dutch} in assuming that the phrase containing the adposition (here \tit{-mee}) can syntactically move to a position higher in the structure. Evidence that \tit{mee} `with' has moved comes from the ungrammaticality of \ref{ex:meeend}. This shows that \tit{mee} `with' cannot be stranded in its base position (see \ref{ex:const3stage}).

@@

So far I showed how \tit{waar-mee} `with what' is derived if all syntactic features form a constituent. Next I address how \tit{waar-mee} `with what' is blocked and \tit{met wat} `with what' appears when the features do not form a proper constituent is derived. An example of a situation in which all features do not form a constituent is given in \ref{ex:noconst}.

@@

This section showed how the instrumental inanimate relative pronoun is realized as \tit{waar-mee} `with what' when all syntactic features form a constituent. It also showed how \tit{met wat} `with what' appears when all features do not form a proper constituent, and other features intervene.

@@

In this paper, I discussed the instrumental \tsc{r}-pronoun and postposition \tit{waar-mee} `with what' in Dutch. The main intuition is that this form that appears when features of an instrumental inanimate object form a proper constituent. The form cannot appear, and \tit{met wat} `with what` surfaces instead, when all features do not form a proper constituent. The described pattern follows from a core assumption in Nanosyntax: phrasal spellout spells out constituents \citep{starke2009}. The distribution of \tit{r}-pronouns and postpositions and prepositions and pronouns follows from regular spellout mechanisms.
